Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-20 paper.tex
Revising related work

Affect recognition is an active field of research. Many
methods have been proposed, some of them requiring the intervention of
the user to fill up questionnaires or forms; these selections remain static until the user changes
the values and are easy to implement, but cannot detect dynamic
changes.  A more dynamic approach requires the use of sensors to capture
affective states in real time.

The context, the environment and the learning
activity determine what kind of sensors can be used. The most common learning
environment is the classroom, a physical space with context to facilitate
learning. But learning is possible in a wide variety of settings, such as
outside-of-school locations and outdoor environments these are sometimes
referred as ubiquitous learning environments, an example of such environments is
the work of Yang \cite{yang2006context} in a context-aware environment, but missing affective
information. There are also virtual learning environments such as the one proposed by Dillenburg
\cite{dillenbourg2002virtual}  where learners can have avatars, and virtual places
where they can play roles and socialize.

In the general context of learning there have been some approaches for affect
recognition. The work of Kapoor y Picard \cite{kapoor2005multimodal} uses a multi modal
approach using sensory information from facial expressions and postural shifts
of the learner combined with information about the learner\'s activity on the
computer; the learning activity was solving a puzzle on a computer. They report
using a multimodal Gaussian Process approach achieving an accuracy of over 86\%.
Sidney, et al., \cite{sidney2005integrating} enhances the intelligent tutor system AutoTutor with
affective recognition using non-intrusive sensory information from facial
expressions, gross body movements, and conversational cues from logs.  The
subject material of that system consisted in lectures of computer
literacy.

Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor} propose the integration of an affective reasoner
with an agent in a virtual environment. In this work an agent called Steve
responds emotionally to other agents and interactive users. The agent simulates
his emotions through different multimedia modes including facial expressions and
speech. Affective reasoning agents were used for training, by putting students
in work related situations. Agents could react to the behavior of students, for
instance if a student was being careless in a task and was in a dangerous
situation, Steve would show distress or fear. Also in a virtual environment the
work of McQuiggan, Robison \& Lester \cite{mcquiggan2010affective} extends this line of research by
investigating the affective transitions that occur throughout narrative-centered
learning experiences. The analysis of affective state transitions in this work
replicated the findings by D’Mello et al. \cite{d2008autotutor} and Baker et al.
\cite{rodrigo2009affective} where
also engagement/flow dominated self-reported affect. For outdoor environments
Shen, Wang \& Shen \cite{shen2009affective} augment a pervasive e-learning platform with affective
recognition.  The results about emotion recognition from physiological signals
achieved a best-case accuracy (86.3\%) for four types of learning emotions.

>>>>

Affect recognition is an active field of research. Many
methods have been proposed, some of them requiring the intervention of
the user to fill up questionnaires or forms; these selections remain static until the user changes
the values and are easy to implement, but cannot detect dynamic
changes.  A more dynamic approach requires the use of sensors to capture
affective states in real time, and that is what we are doing in this
paper.

The actual sensors are determined by the context, the environment and the learning
activity determine. The most common learning
environment is the classroom, a physical space that provides a context to facilitate
learning. But learning is possible in a wide variety of settings, such as
out-of-school locations and outdoor environments which are sometimes
referred to as ubiquitous learning environments. An analysis of such
environments is done by Yang \cite{yang2006context} in a context-aware
environment, but his work misses affective information. There are also
virtual learning environments such as the one proposed by Dillenburg 
\cite{dillenbourg2002virtual}  where learners can have avatars, and virtual places
where they can play roles and socialize. In general, the importance of
affect in learning environments has been recognized at a later stage
of research.

For instance, the work of Kapoor y Picard \cite{kapoor2005multimodal}
uses a multimodal 
approach with sensory information from facial expressions and postural shifts
of the learner combined with information about the learner\'s activity on the
computer; the learning activity was solving a puzzle on a computer. They report
using a multimodal Gaussian Process approach achieving an accuracy of over 86\%.
Sidney, et al., \cite{sidney2005integrating} enhances the intelligent tutor system AutoTutor with
affective recognition using non-intrusive sensory information from facial
expressions, gross body movements, and conversational cues from logs.  The
subject material of that system consisted in lectures of computer
literacy. In this early stage of research, the importance of privacy
and non-intrusiveness is set aside.

Some other works try to introduce affective computing features into
interactive learning environments, without explicitly detecting of
using the student's affective state. For instance, Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor}
propose the integration of an affective reasoner 
with an agent in a virtual environment. In this work an agent called Steve
responds emotionally to other agents and interactive users. The agent simulates
his emotions through different multimedia modes including facial expressions and
speech. Affective reasoning agents were used for training, by putting students
in work related situations. Agents could react to the behavior of students; for
instance if a student was being careless in a task and was in a dangerous
situation, Steve would show distress or fear. Also in a virtual environment the
work of McQuiggan, Robison \& Lester \cite{mcquiggan2010affective} extends this line of research by
investigating the affective transitions that occur throughout narrative-centered
learning experiences. The analysis of affective state transitions in this work
replicated the findings by D’Mello et al. \cite{d2008autotutor} and Baker et al.
\cite{rodrigo2009affective} where
also engagement/flow dominated self-reported affect. For outdoor environments
Shen, Wang \& Shen \cite{shen2009affective} augment a pervasive e-learning platform with affective
recognition.  The results about emotion recognition from physiological signals
achieved a best-case accuracy (86.3\%) for four types of learning emotions.
